On Misuses of the Kolmogorov--Smirnov Test for One-Sample Goodness-of-Fit

### 2023-05-09 manuscript/manuscript.tex
minor

The standard one-sample KS test applies to independent data with a continuous
hypothesized distribution that is completely specified. In practice, however, it
has often been applied without realizing that one or more of these assumptions
do not hold. For example, \citet{noether1963note} showed the conservativeness of
the KS test when applied to discontinuous distributions. The null distribution
of the KS statistic is no longer distribution free. Computing the
exact and asymptotic distribution of $D_n$ is challenging. Fortunately, the null
distribution of the KS statistic for discontinuous distributions has been
efficiently addressed by \citet{dimitrova2020computing} with a
companion R package \textsf{KSgeneral}. Although a common misuse of KS test, the
issue with discontinuous data is not our focus.

When the hypothesized distribution~$F$ contains unspecified parameters, as is
the case in most goodness-of-fit test settings, the standard KS test is not
applicable. \citet{steinskog2007cautionary} ``discovers'' the change in power
when using fitted parameters and stresses caution in using the KS test in
such ways. In fact, using fitted parameters in place of the true parameters in
KS test has been long known to yield extremely conservative results
\citep[e.g.,][]{lilliefors1967kolmogorov}. This problem can be solved by
parametric bootstrap \citep{efron1985bootstrap, hall1991two}, where
bootstrap samples of the test statistics are constructed from samples
generated from the fitted hypothesized distribution. A nonparametric bootstrap
solution is not trivial because a nonparametric bootstrap sample of the
observed data has ties, which would not happen for continuous distributions.
\citet{babu2004goodness} derived the bias of the standard nonparametric 
bootstrap and
showed how to correct it. They further noted that both parametric and
nonparametric procedures lead to correct asymptotic levels.

@@

The consequence of using the wrong null distribution for the KS statistic $D_n$
in~\eqref{eq:ks_fitted} can be illustrated through a simple simulation study. A
random sample $X_1, \ldots, X_n$ was generated from a normal
distribution with both mean and variance parameters set equal to 8 ($N(8,8)$), 
and sample size $n = 200$. The maximum likelihood estimates of the mean and 
variance were used as $\hat\theta_n$. The p-value of the test
statistic~$D_n$ in Equation~\eqref{eq:ks_fitted} was obtained by naively calling 
the \texttt{ks.test()} function in~R with the fitted normal distribution as the
hypothesized distribution. That is, the hypothesized distribution was
$N(\bar X, s^2)$ where $\bar X$ is the sample mean and $s^2$ is the sample
variance (with $n$ in the denominator). This experiment was repeated 1000 times
and the probability-probability plot (PP-plot) of the 1000 p-values are 
displayed in the top left plot of Figure~\ref{fig:pp_f}. If the test were
valid, the p-values would be uniformly distributed in the unit
interval and the points in the PP plot would fall along the diagonal line. 
Clearly, the distribution of the 1000 p-values are very different from what
one would expect to see from 1000 draws from the standard uniform
distribution.

@@

We similarly considered random samples $X_1, \ldots, X_n$ 
generated from a gamma distribution with shape and scale parameters 8 and 1, 
respectively ($\Gamma(8,1)$), and sample size $n=200$. These parameter values 
were selected so that mean and variance of the normal and gamma distributions 
in the simulations are the same. Likewise, the p-value of the test 
statistic~$D_n$ in~\eqref{eq:ks_fitted} was obtained by naively calling the
\texttt{ks.test()} function in~R with the fitted gamma distribution as the
hypothesized distribution using the shape and scale maximum likelihood 
estimates. The PP plot of the p-values from 1000 replications of this 
experiment are displayed in the bottom left plot of Figure~\ref{fig:pp_f}. The
p-values are again non-uniformly distributed due to the use of the 
estimated parameters when specifying the null distribution.

@@

Nonparametric bootstrap can also be used to approximate the null distribution
of the test statistic in~\eqref{eq:ks_fitted} except that a bias correction is 
needed \citep{babu2004goodness}; see details of the procedure in
in the Appendix. The results from nonparametric bootstrap are similar to those
from the parametric bootstrap and, hence, are omitted.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=\textwidth]{pp_s.pdf}
  \caption{PP-plots for the p-values obtained from KS test with serial
    dependence ignored and sample size $n = 200$ based on 1000
    replicates. The data have marginal distributions $N(8, 8)$ and
    $\Gamma(8, 1)$, with an AR(1) dependence structure on the standard normal
    scale. The AR coefficients were $\psi \in \{-0.4, -0.2,  0.2,  0.4\}$.
  }
  \label{fig:pp_s}
\end{figure}

The invalidity of the standard statistic for serially dependent data can be
illustrated by a simple simulation study. Consider a first-order autoregressive
or AR(1) model with marginal normal (or gamma) distribution. For each AR
coefficient $\psi \in \{-0.4, -0.2, 0.2, 0.4\}$, we generated 1000
series of length 200. For each series, we applied the KS statistic
in~\eqref{eq:ks_standard} to test $H_0$ that the $X_i$'s follow $N(8, 8)$ (or 
$\Gamma(8,1)$) marginally. The PP-plots of the p-values from $1000$ replicates 
for all the $\psi$ values are displayed in Figure~\ref{fig:pp_s}. As 
$\phi$ diverges from 0, the p-values look less likely to be generated from a 
standard uniform distributions. Positive $\psi$ values led to p-values with a 
distribution function higher than the standard uniform distribution function as 
there were more smaller p-values than there should be, and hence, 
liberal tests. Negative $\psi$ values led to p-values with a 
distribution function lower than the standard uniform distribution as there 
were more larger p-values than there should be, and hence, conservative tests.

@@

The semiparametric bootstrap procedure was evaluated in a simulation study. We
considered two marginal distributions: $N(8,8)$ and $\Gamma(8,1)$. Three
dependence structures were used to link variables generated from each marginal
distribution; on the standard normal scale, they are
ARMA(1, 1) with AR coefficient 0.5 and MA coefficient 0.3;
AR(2) with coefficients $(0.6, 0.2)$; and
MA(1)) with coefficient 0.8.
For each dependence structure and marginal distribution combination, we
generated 1000 series of length 200.  For each series, we applied the
semiparametric bootstrap procedure with serial
dependence accounted for through a working ARMA model to test $H_0$ that the 
$X_i$'s marginally follow their true marginal distribution ($N(8, 8)$ or 
$\Gamma(8,1)$). Figure~\ref{fig:pp_ss} displays the PP-plots of the 1000 
p-values for each dependence structure and marginal distribution combination.
All plots within the figure show the points falling along the diagonal lines, 
implying that the p-values are coming from standard uniform distributions and 
that the semiparametric bootstrap procedure is an effective remedy in all cases
considered.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=.75\textwidth]{pp_s_f.pdf}
  \includegraphics[width=.75\textwidth]{pp_ssf.pdf}
  \caption{PP-plots for the p-values obtained from KS test without (top) and 
	  with (bottom)
    serial dependence accounted for and unspecified parameters and sample size
    $n = 200$ based on
    1000 replicates. The data have marginal distributions $N(8, 8)$ and
    $\Gamma(8, 1)$. The dependence structure on the standard normal scale was
    characterized by ARMA(1, 1) model, AR(2) model, and MA(1) model.
  }
  \label{fig:pp_ssf}
\end{figure}

@@

Finally, we report a small study of the power of the one-sample KS test with
unspecified parameters and serially dependent data. Using the three dependence
models characterized by ARMA(1, 1), AR(2), and MA(1) in the simulation settings
in the last section, we generated serially
dependent data with marginal distributions $N(8, 8)$ left truncated by zero and
$\Gamma(8, 1)$. 
Then, for data from $\Gamma(8, 1)$, we tested $H_0$ that the
data has marginally a normal distribution; for data from $N(8, 8)$ left
truncated by zero, we tested $H_0$ that the data has marginally a gamma
distribution;  In both cases, no parameters were specified. Two sample sizes
were considered, $n \in \{100, 200\}$. Table~\ref{tab:power} summarizes the
power of rejecting the null hypothesis at significance level~0.05 from
1000~replicates. The tests
based on a working copula have substantial power in rejecting the null
hypothesis and the power increases as sample size increases. The power from the
naive tests where the serial dependence is ignored, although the unspecified
parameters are accounted for (i.e., procedure from 
Section~\ref{sec:dependence}), appear to 
have higher power than the proposed tests, but this is not reliable since they 
do not hold their sizes as demonstrated in the Figure~\ref{fig:pp_ssf}.

@@

The three kinds of misuses of the KS test for one-sample goodness-of-fit that we
demonstrated here are commonly seen in practice. With the aid of computing tools
such as R, they can be easily demonstrated in classroom teaching. Practitioners
need to be aware of the assumptions of the standard one-sample KS test that the
hypothesized distribution is completely specified and that the data are
independent. When these assumptions are not met, naive applications of the test
are no longer accurate and remedies must be performed. In the
case of fitted parameters, parametric and nonparametric bootstrap can restore
the size of the test. The nonparametric version has long been developed too but
is lesser known \citep{babu2004goodness}. In the case of serially dependent data,
our semiparametric bootstrap procedure with a working dependence structure
controlled by an automatically selected ARMA model restore the size of the test
for a wide range of dependence structures.
When both assumptions are violated, i.e., where the data has serial 
dependence and the parameters must be fitted, a combination of the two
individual remedies shows good results.

@@

Using the same notations in the text, the nonparametric bootstrap procedure of
\citet{babu2004goodness} is summarized as follows
\citep{babu2004goodness}.
\begin{enumerate}
\item
  Draw a random sample $X_1^*,...,X_n^*$ from the empirical distribution $F_n$
  with replacement
\item
  Fit $F_\theta$ to $X_1^*,...,X_n^*$ and obtain estimate 
	$\hat\theta_n^*$ of $\theta$.
\item
  Obtain the empirical distribution function $F_n^*$ of
  $X_1^*, \ldots,  X_n^*$. 
\item
  Calculate bootstrap KS statistic
  \[
    D_n^* = \sup_x | \sqrt{n}\left(F_n^* (x)- F_{\hat\theta_n^*}(x)\right) - B_n(x) |.
  \]
  where $B_{n}(x) = \sqrt{n}(F_{n}(x) - F_{\hat\theta_n}(x))$ is the known
  bias term \citep{babu2004goodness}.
\item
  Repeat the previous steps a large number $B$ times and use the empirical
  distribution of $D_n^*$ to approximate the null distribution of the observed
  statistic. 
\end{enumerate}
The p-value of $D_n$ is, again, approximated by the proportion of the $D_n^*$
statistics that are greater than or equal to~$D_n$. Note that the step~1 and
step~4 are different from the parametric bootstrap version.

>>>>

The standard one-sample KS test applies to independent data with a continuous
hypothesized distribution that is completely specified. In practice, however, it
has often been applied without realizing that one or more of these assumptions
do not hold. For example, \citet{noether1963note} showed the conservativeness of
the KS test when applied to discontinuous distributions. The null distribution
of the KS statistic is no longer distribution free. Computing the
exact and asymptotic distribution of $D_n$ is challenging. Fortunately, the null
distribution of the KS statistic for discontinuous distributions has been
efficiently addressed by \citet{dimitrova2020computing} with a
companion R package \textsf{KSgeneral}. Although a common misuse of the KS test, 
the issue with discontinuous data is not our focus.

When the hypothesized distribution~$F$ contains unspecified parameters, as is
the case in most goodness-of-fit test settings, the standard KS test is not
applicable. \citet{steinskog2007cautionary} ``discovers'' the change in power
when using fitted parameters and stresses caution in using the KS test in
such ways. In fact, using fitted parameters in place of the true parameters in
the KS test has been long known to yield extremely conservative results
\citep[e.g.,][]{lilliefors1967kolmogorov}. This problem can be solved by
parametric bootstrap \citep{efron1985bootstrap, hall1991two}, where
bootstrap samples of the test statistics are constructed from samples
generated from the fitted hypothesized distribution. A nonparametric bootstrap
solution is not trivial because a nonparametric bootstrap sample of the
observed data has ties, which would not happen for continuous distributions.
\citet{babu2004goodness} derived the bias of the standard nonparametric 
bootstrap and
showed how to correct it. They further noted that both parametric and
nonparametric procedures lead to correct asymptotic levels.

@@

The consequence of using the wrong null distribution for the KS statistic $D_n$
in~\eqref{eq:ks_fitted} can be illustrated through a simple simulation study. A
random sample $X_1, \ldots, X_n$ was generated from a normal
distribution with both mean and variance parameters set equal to 8 ($N(8,8)$), 
and sample size $n = 200$. The maximum likelihood estimates of the mean and 
variance were used as $\hat\theta_n$. The p-value of the test
statistic~$D_n$ in Equation~\eqref{eq:ks_fitted} was obtained by naively calling 
the \texttt{ks.test()} function in~R with the fitted normal distribution as the
hypothesized distribution. That is, the hypothesized distribution was
$N(\bar X, s^2)$ where $\bar X$ is the sample mean and $s^2$ is the sample
variance (with $n$ in the denominator). This experiment was repeated 1000 times
and the probability-probability plot (PP-plot) of the 1000 p-values are 
displayed in the top left plot of Figure~\ref{fig:pp_f}. If the test were
valid, the p-values would be uniformly distributed in the unit
interval and the points in the PP-plot would fall along the diagonal line. 
Clearly, the distribution of the 1000 p-values are very different from what
one would expect to see from 1000 draws from the standard uniform
distribution.

@@

We similarly considered random samples $X_1, \ldots, X_n$ 
generated from a gamma distribution with shape and scale parameters 8 and 1, 
respectively ($\Gamma(8,1)$), and sample size $n=200$. These parameter values 
were selected so that mean and variance of the normal and gamma distributions 
in the simulations are the same. Likewise, the p-value of the test 
statistic~$D_n$ in~\eqref{eq:ks_fitted} was obtained by naively calling the
\texttt{ks.test()} function in~R with the fitted gamma distribution as the
hypothesized distribution using the shape and scale maximum likelihood 
estimates. The PP-plot of the p-values from 1000 replications of this 
experiment are displayed in the bottom left plot of Figure~\ref{fig:pp_f}. The
p-values are again non-uniformly distributed due to the use of the 
estimated parameters when specifying the null distribution.

@@

Nonparametric bootstrap can also be used to approximate the null distribution
of the test statistic in~\eqref{eq:ks_fitted} except that a bias correction is 
needed \citep{babu2004goodness}; see details of the procedure
in the Appendix. The results from nonparametric bootstrap are similar to those
from the parametric bootstrap and, hence, are omitted.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=\textwidth]{pp_s.pdf}
  \caption{PP-plots for the p-values obtained from KS test with serial
    dependence ignored and sample size $n = 200$ based on 1000
    replicates. The data have marginal distributions $N(8, 8)$ and
    $\Gamma(8, 1)$, with an AR(1) dependence structure on the standard normal
    scale. The AR coefficients are $\psi \in \{-0.4, -0.2,  0.2,  0.4\}$.
  }
  \label{fig:pp_s}
\end{figure}

The invalidity of the standard statistic for serially dependent data can be
illustrated by a simple simulation study. Consider a first-order autoregressive
or AR(1) model with marginal normal (or gamma) distribution. For each AR
coefficient $\psi \in \{-0.4, -0.2, 0.2, 0.4\}$, we generated 1000
series of length 200. For each series, we applied the KS statistic
in~\eqref{eq:ks_standard} to test $H_0$ that the $X_i$'s follow $N(8, 8)$ (or 
$\Gamma(8,1)$) marginally. The PP-plots of the p-values from $1000$ replicates 
for each $\psi$ value are displayed in Figure~\ref{fig:pp_s}. As 
$\psi$ diverges from 0, the p-values look less likely to be generated from a 
standard uniform distributions. Positive $\psi$ values led to p-values with a 
distribution function higher than the standard uniform distribution function as 
there were more smaller p-values than there should be, and hence, 
liberal tests. Negative $\psi$ values led to p-values with a 
distribution function lower than the standard uniform distribution as there 
were more larger p-values than there should be, and hence, conservative tests.

@@

The semiparametric bootstrap procedure was evaluated in a simulation study. We
considered two marginal distributions: $N(8,8)$ and $\Gamma(8,1)$. Three
dependence structures were used to link variables generated from each marginal
distribution; on the standard normal scale, they are
ARMA(1, 1) with AR coefficient 0.5 and MA coefficient 0.3;
AR(2) with coefficients $(0.6, 0.2)$; and
MA(1) with coefficient 0.8.
For each dependence structure and marginal distribution combination, we
generated 1000 series of length 200.  For each series, we applied the
semiparametric bootstrap procedure with serial
dependence accounted for through a working ARMA model to test $H_0$ that the 
$X_i$'s marginally follow their true marginal distribution ($N(8, 8)$ or 
$\Gamma(8,1)$). Figure~\ref{fig:pp_ss} displays the PP-plots of the 1000 
p-values for each dependence structure and marginal distribution combination.
All plots within the figure show the points falling along the diagonal lines, 
implying that the p-values are coming from standard uniform distributions and 
that the semiparametric bootstrap procedure is an effective remedy in all cases
considered.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=.75\textwidth]{pp_s_f.pdf}
  \includegraphics[width=.75\textwidth]{pp_ssf.pdf}
  \caption{PP-plots for the p-values obtained from KS test without (top) and 
	  with (bottom)
    serial dependence accounted for and unspecified parameters; sample size
    $n = 200$ based on
    1000 replicates. The data have marginal distributions $N(8, 8)$ and
    $\Gamma(8, 1)$. The dependence structure on the standard normal scale was
    characterized by ARMA(1, 1) model, AR(2) model, and MA(1) model.
  }
  \label{fig:pp_ssf}
\end{figure}

@@

Finally, we report a small study of the power of the one-sample KS test with
unspecified parameters and serially dependent data. Using the three dependence
models characterized by ARMA(1, 1), AR(2), and MA(1) in the simulation settings
in the last section, we generated serially
dependent data with marginal distributions $N(8, 8)$ left truncated by zero and
$\Gamma(8, 1)$. 
Then, for data from $\Gamma(8, 1)$, we tested $H_0$ that the
data has marginally a normal distribution; for data from $N(8, 8)$ left
truncated by zero, we tested $H_0$ that the data has marginally a gamma
distribution.  In both cases, no parameters were specified. Two sample sizes
were considered, $n \in \{100, 200\}$. Table~\ref{tab:power} summarizes the
power of rejecting the null hypothesis at significance level~0.05 from
1000~replicates. The tests
based on a working copula have substantial power in rejecting the null
hypothesis and the power increases as sample size increases. The power from the
naive tests where the serial dependence is ignored, although the unspecified
parameters are accounted for (i.e., procedure from 
Section~\ref{sec:fitted}), appear to 
have higher power than the proposed tests, but these are not reliable since they 
do not hold their sizes as demonstrated in the Figure~\ref{fig:pp_ssf}.

@@

The three kinds of misuses of the KS test for one-sample goodness-of-fit that we
demonstrated here are commonly seen in practice. With the aid of computing tools
such as R, they can be easily demonstrated in classroom teaching. Practitioners
need to be aware of the assumptions of the standard one-sample KS test that the
hypothesized distribution is completely specified and that the data are
independent. When these assumptions are not met, naive applications of the test
are no longer accurate and remedies must be performed. In the
case of fitted parameters, parametric and nonparametric bootstrap can restore
the size of the test. The nonparametric version has long been developed too but
is lesser known \citep{babu2004goodness}. In the case of serially dependent data,
our semiparametric bootstrap procedure with a working dependence structure
controlled by an automatically selected ARMA model restores the size of the test
for a wide range of dependence structures.
When both assumptions are violated, i.e., where the data has serial 
dependence and the parameters must be fitted, a combination of the two
individual remedies shows good results.

@@

Using the same notations in the text, the nonparametric bootstrap procedure of
\citet{babu2004goodness} is summarized as follows
\citep{babu2004goodness}.
\begin{enumerate}
\item
  Draw a random sample $X_1^*,...,X_n^*$ from the empirical distribution $F_n$
  with replacement.
\item
  Fit $F_\theta$ to $X_1^*,...,X_n^*$ and obtain estimate 
	$\hat\theta_n^*$ of $\theta$.
\item
  Obtain the empirical distribution function $F_n^*$ of
  $X_1^*, \ldots,  X_n^*$. 
\item
  Calculate bootstrap KS statistic
  \[
    D_n^* = \sup_x | \sqrt{n}\left(F_n^* (x)- F_{\hat\theta_n^*}(x)\right) - B_n(x) |.
  \]
  where $B_{n}(x) = \sqrt{n}(F_{n}(x) - F_{\hat\theta_n}(x))$ is the known
  bias term \citep{babu2004goodness}.
\item
  Repeat the previous steps a large number $B$ times and use the empirical
  distribution of $D_n^*$ to approximate the null distribution of the observed
  statistic. 
\end{enumerate}
The p-value of $D_n$ is, again, approximated by the proportion of the $D_n^*$
statistics that are greater than or equal to~$D_n$. Note that the step~1 and
step~4 are different from the parametric bootstrap version.
